---
title: "Walmart M5 Sales Forecasting: Time Series Analysis"
author: "Suin Kim"
output: html_document
---

## Project Overview

This project applies time series forecasting methods to the **M5 Forecasting Competition** dataset — a real-world retail sales dataset from Walmart covering 3,049 products across 10 stores in 3 US states (California, Texas, Wisconsin) from 2011 to 2016.

**Business Goal:** Forecast daily aggregate sales to support inventory planning and demand management. Accurate sales forecasts reduce overstock and stockout costs — critical operational challenges for large retailers.

**Models compared:**
- ARIMA (AutoRegressive Integrated Moving Average)
- SARIMA (Seasonal ARIMA)
- TBATS (Trigonometric seasonality, Box-Cox transformation, ARMA errors, Trend, Seasonal)

**Data source:** [M5 Forecasting Accuracy — Kaggle](https://www.kaggle.com/competitions/m5-forecasting-accuracy)

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyr)
library(forecast)
library(astsa)
library(Metrics)
library(lubridate)

# Dataset files needed in working directory:
# - sales_train_validation.csv
# - calendar.csv
# Download from: https://www.kaggle.com/competitions/m5-forecasting-accuracy
```

---

## 2. Data Loading & Preparation

### 2.1 Dataset Description

The M5 dataset has two key files:

- **sales_train_validation.csv**: Each row is one product-store combination. Columns `d_1` through `d_1941` represent daily unit sales for each calendar day.
- **calendar.csv**: Maps day codes (`d_1`, `d_2`, ...) to actual dates, with additional features for events (holidays, sporting events) and SNAP benefit days by state.

```{r load-data, message=FALSE, warning=FALSE}
setwd("/Users/suinkim/Desktop/Time_Series")
sales_train_validation <- fread("sales_train_validation.csv")
calendar <- fread("calendar.csv")
calendar <- as.data.table(calendar)

cat("Sales data dimensions:", dim(sales_train_validation), "\n")
cat("Calendar dimensions:", dim(calendar), "\n")
cat("Date range:", as.character(min(calendar$date)), "to", as.character(max(calendar$date)), "\n")
cat("Number of unique products:", length(unique(sales_train_validation$item_id)), "\n")
cat("Number of stores:", length(unique(sales_train_validation$store_id)), "\n")
cat("States covered:", paste(unique(sales_train_validation$state_id), collapse = ", "), "\n")
cat("Product categories:", paste(unique(sales_train_validation$cat_id), collapse = ", "), "\n")
```

### 2.2 Pivot & Merge with Calendar

```{r pivot-merge, message=FALSE, warning=FALSE}
# Pivot from wide format (d_1, d_2, ...) to long format (day, sales)
# This transforms 1941 columns into rows for easier manipulation
sales_pivot <- sales_train_validation %>%
  pivot_longer(
    cols       = starts_with("d_"),
    names_to   = "day",
    values_to  = "sales"
  )

# Convert to data.table for faster processing on large dataset
sales_pivot <- as.data.table(sales_pivot)

# Merge with calendar to attach actual dates and event info
setkey(calendar, d)
sales_with_calendar <- merge(
  sales_pivot,
  calendar,
  by.x   = "day",
  by.y   = "d",
  all.x  = TRUE
)

# Filter to 2014-2015 for modeling
# These two years provide enough data for seasonal patterns
# while keeping computation manageable
sales_with_calendar <- sales_with_calendar[year %in% c(2014, 2015)]
sales_with_calendar <- sales_with_calendar[, .(cat_id, state_id, sales, date, weekday)]

cat("Filtered data dimensions:", dim(sales_with_calendar), "\n")
cat("Date range after filtering:", as.character(min(sales_with_calendar$date)),
    "to", as.character(max(sales_with_calendar$date)), "\n")
```

---

## 3. Exploratory Data Analysis

### 3.1 Daily Total Sales & Outlier Removal

Before any exploration, we first aggregate to daily totals and immediately remove the Christmas Day structural outlier. December 25th records near-zero sales because Walmart stores are closed — this is not random noise but a known structural zero that would distort all downstream analysis and models.

```{r eda-daily-raw, fig.width=12, fig.height=5}
# Aggregate to total daily sales across all products and stores
d_sales <- sales_with_calendar[, .(total_sales = sum(sales)), by = date]
d_sales <- d_sales[order(date)]

# Show raw data first — Christmas outliers clearly visible as sharp vertical drops
ggplot(d_sales, aes(x = date, y = total_sales)) +
  geom_line(color = "steelblue") +
  labs(
    title    = "Daily Total Sales: 2014–2015 (Raw)",
    subtitle = "Two sharp drops in December = Christmas closures (structural zeros — not random noise)",
    x        = "Date",
    y        = "Total Units Sold"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

```{r outlier, fig.width=12, fig.height=5}
# Identify and remove Christmas Day outliers
low_sales_dates <- d_sales[lubridate::month(date) == 12 & lubridate::day(date) == 25]

cat("Christmas Day sales removed:\n")
print(low_sales_dates)

# Remove Christmas dates — all remaining EDA and modeling uses this clean series
d_sales <- d_sales[!(date %in% low_sales_dates$date)]

# Also remove from sales_with_calendar so category/state charts are consistent
sales_with_calendar <- sales_with_calendar[!(lubridate::month(date) == 12 & lubridate::day(date) == 25)]

ggplot(d_sales, aes(x = date, y = total_sales)) +
  geom_line(color = "steelblue") +
  labs(
    title    = "Daily Total Sales: 2014–2015 (Christmas Removed)",
    subtitle = "Weekly seasonality now clearly visible — used for all subsequent analysis",
    x        = "Date",
    y        = "Total Units Sold"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

### 3.2 Sales by Category

```{r eda-category, fig.width=12, fig.height=5}
# Examine sales distribution across product categories (Christmas already removed)
# FOODS dominates — justifies aggregate forecasting as a starting point
cat_sales <- sales_with_calendar[, .(total_sales = sum(sales)), by = .(date, cat_id)]

ggplot(cat_sales, aes(x = date, y = total_sales, color = cat_id)) +
  geom_line(alpha = 0.8) +
  labs(
    title  = "Daily Sales by Category: 2014–2015",
    x      = "Date",
    y      = "Total Units Sold",
    color  = "Category"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

### 3.3 Sales by State

```{r eda-state, fig.width=12, fig.height=5}
# California consistently leads in sales volume across both years
state_sales <- sales_with_calendar[, .(total_sales = sum(sales)), by = .(date, state_id)]

ggplot(state_sales, aes(x = date, y = total_sales, color = state_id)) +
  geom_line(alpha = 0.8) +
  labs(
    title  = "Daily Sales by State: 2014–2015",
    x      = "Date",
    y      = "Total Units Sold",
    color  = "State"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

### 3.4 Day-of-Week Patterns

```{r eda-weekday, fig.width=10, fig.height=5}
# Weekly seasonality check — retail sales typically peak on weekends
# This confirms s=7 is the right seasonal period for SARIMA and TBATS
weekday_sales <- sales_with_calendar[, .(avg_sales = mean(sales)), by = weekday]
weekday_order <- c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")
weekday_sales$weekday <- factor(weekday_sales$weekday, levels = weekday_order)

ggplot(weekday_sales, aes(x = weekday, y = avg_sales, fill = weekday)) +
  geom_bar(stat = "identity") +
  labs(
    title    = "Average Sales by Day of Week",
    subtitle = "Confirms strong weekly seasonality — critical for model selection",
    x        = "Day of Week",
    y        = "Average Units Sold"
  ) +
  theme_minimal() +
  theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "none")
```

### 3.5 Yearly Mean and Variance

```{r eda-stationarity-check, fig.width=10, fig.height=6}
# Check whether mean and variance are stable over time
# Changing mean/variance = non-stationary -> differencing needed
d_sales$year <- as.numeric(format(d_sales$date, "%Y"))
stats_by_year <- d_sales[, .(mean_sales = mean(total_sales),
                               var_sales  = var(total_sales)), by = year]

cat("Yearly statistics:\n")
print(stats_by_year)

par(mfrow = c(2, 1), mar = c(4, 5, 3, 2))
plot(stats_by_year$year, stats_by_year$mean_sales,
     type = "b", col = "steelblue", pch = 19,
     main = "Yearly Mean Sales", xlab = "Year", ylab = "Mean Sales")
plot(stats_by_year$year, stats_by_year$var_sales,
     type = "b", col = "tomato", pch = 19,
     main = "Yearly Variance of Sales", xlab = "Year", ylab = "Variance")
par(mfrow = c(1, 1))
```

Both mean and variance shift across years — confirming the series is **non-stationary** and requires differencing before ARIMA modeling.

---

## 4. Stationarity Testing

### 4.1 Augmented Dickey-Fuller Test

The **Augmented Dickey-Fuller (ADF) test** formally tests for a unit root (non-stationarity).

$$H_0: \text{Series has a unit root (non-stationary)}$$
$$H_1: \text{Series is stationary}$$

If $p > 0.05$: fail to reject $H_0$ — series is non-stationary, differencing required.

```{r adf-test}
library(tseries)

cat("=== ADF Test on Raw Sales ===\n")
adf_raw <- adf.test(d_sales$total_sales)
print(adf_raw)

if (adf_raw$p.value > 0.05) {
  cat("-> Non-stationary. Differencing required.\n")
} else {
  cat("-> Stationary. No differencing needed.\n")
}
```

### 4.2 First-Order Differencing

```{r differencing, fig.width=12, fig.height=5}
# First-order differencing: removes trend by computing day-over-day change
# Δy_t = y_t - y_{t-1}
d_sales_diff <- data.frame(
  date       = d_sales$date[-1],
  diff_sales = diff(d_sales$total_sales, lag = 1)
)

ggplot(d_sales_diff, aes(x = date, y = diff_sales)) +
  geom_line(color = "steelblue") +
  labs(
    title    = "First-Order Differenced Daily Sales",
    subtitle = "Δy_t = y_t - y_{t-1} — trend removed, series oscillates around zero",
    x        = "Date",
    y        = "Differenced Sales"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

# Confirm stationarity after differencing
cat("\n=== ADF Test on Differenced Sales ===\n")
adf_diff <- adf.test(d_sales_diff$diff_sales)
print(adf_diff)

if (adf_diff$p.value <= 0.05) {
  cat("-> Stationary after differencing. Proceed with ARIMA.\n")
}
```

---

## 5. ACF & PACF Analysis

ACF and PACF plots are used to identify the AR order (p) and MA order (q) for the ARIMA model.

**Interpretation rules:**
- **ACF** — significant lags beyond lag 0 suggest MA terms (q)
- **PACF** — significant lags beyond lag 0 suggest AR terms (p)
- Exponential decay in ACF with sharp PACF cutoff → AR model
- Sharp ACF cutoff with exponential PACF decay → MA model

```{r acf-pacf, fig.width=12, fig.height=6}
par(mfrow = c(1, 2))
Acf(d_sales_diff$diff_sales,
    main = "ACF — Differenced Sales",
    lag.max = 40)
Pacf(d_sales_diff$diff_sales,
     main = "PACF — Differenced Sales",
     lag.max = 40)
par(mfrow = c(1, 1))
```

**Reading the plots:**
- ACF shows significant spikes at lags 1, 7 — suggesting MA(1) and weekly seasonal component
- PACF shows significant spikes at lags 1–3 — suggesting AR(2) or AR(3) as starting point
- Spike at lag 7 in both ACF/PACF confirms 7-day weekly seasonality

---

## 6. Data Smoothing

```{r smoothing, fig.width=12, fig.height=6}
# Moving averages smooth short-term fluctuations to reveal underlying trends
# 7-day MA captures weekly cycle; 30-day MA captures monthly trend

# Remove extreme outliers before smoothing (threshold: 10,000 units change/day)
threshold <- 10000
d_sales_diff_cleaned <- d_sales_diff[abs(d_sales_diff$diff_sales) < threshold, ]

ma7  <- stats::filter(d_sales_diff_cleaned$diff_sales, sides = 2, rep(1, 7) / 7)
ma30 <- stats::filter(d_sales_diff_cleaned$diff_sales, sides = 2, rep(1, 30) / 30)

plot(d_sales_diff_cleaned$date, d_sales_diff_cleaned$diff_sales,
     type = "l", col = "grey60",
     main = "Data Smoothing: 7-Day and 30-Day Moving Averages",
     xlab = "Date", ylab = "Differenced Sales")
lines(d_sales_diff_cleaned$date, ma7,  col = "steelblue", lwd = 2)
lines(d_sales_diff_cleaned$date, ma30, col = "tomato",    lwd = 2, lty = 2)
legend("topright",
       legend = c("Original", "7-Day MA", "30-Day MA"),
       col    = c("grey60", "steelblue", "tomato"),
       lty    = c(1, 1, 2), lwd = 1, cex = 0.8)
```

---

## 7. Seasonal Decomposition

```{r decomposition, fig.width=12, fig.height=8}
# Decompose the raw series into trend, seasonal, and residual components
# Using period = 7 for weekly seasonality
# This gives us a clearer view of the underlying structure before modeling

sales_ts <- ts(d_sales$total_sales, frequency = 7)
decomp   <- stl(sales_ts, s.window = "periodic")
plot(decomp,
     main = "STL Decomposition: Trend + Weekly Seasonality + Residuals")
```

**Decomposition findings:**
- **Trend**: Gradual upward trend through 2014-2015
- **Seasonal**: Strong, consistent 7-day cycle — weekend peaks, mid-week troughs
- **Remainder**: Random noise after trend and seasonality are removed

---

## 8. Train / Test Split

```{r train-test}
# 90/10 split — use 90% of data for training, hold out last 10% for evaluation
n          <- nrow(d_sales_diff)
train_size <- floor(0.9 * n)

train_data <- d_sales_diff[1:train_size, ]
test_data  <- d_sales_diff[(train_size + 1):n, ]

cat("Training observations:", nrow(train_data), "\n")
cat("Testing observations: ", nrow(test_data), "\n")
cat("Train period:", as.character(min(train_data$date)),
    "to", as.character(max(train_data$date)), "\n")
cat("Test period: ", as.character(min(test_data$date)),
    "to", as.character(max(test_data$date)), "\n")
```

---

## 9. Model 1: ARIMA

### 9.1 Model Specification

The **ARIMA(p, d, q)** model combines three components:

$$\Phi(B)(1-B)^d y_t = \Theta(B)\epsilon_t$$

Where:
- $p$ = autoregressive order (how many past values predict the current value)
- $d$ = degree of differencing (how many times the series is differenced to achieve stationarity)
- $q$ = moving average order (how many past forecast errors predict the current value)
- $B$ = backshift operator: $By_t = y_{t-1}$
- $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$ = white noise

**For our data:** $d = 1$ (first-order differencing confirmed stationary), so we model on the differenced series with $d = 0$ (pre-differenced input).

### 9.2 AIC Grid Search for Optimal p and q

```{r arima-grid, warning=FALSE}
# Grid search over p in {0,...,4} and q in {0,...,4}
# Select model with lowest AIC (Akaike Information Criterion)
# AIC penalizes model complexity: AIC = 2k - 2ln(L)
# where k = number of parameters, L = likelihood

p_range  <- 0:4
q_range  <- 0:4
best_aic <- Inf
best_p   <- 0
best_q   <- 0
best_model <- NULL

for (p in p_range) {
  for (q in q_range) {
    model <- try(Arima(train_data$diff_sales, order = c(p, 0, q)), silent = TRUE)
    if (!inherits(model, "try-error")) {
      current_aic <- AIC(model)
      if (current_aic < best_aic) {
        best_aic   <- current_aic
        best_model <- model
        best_p     <- p
        best_q     <- q
      }
    }
  }
}

cat("Best ARIMA order: ARIMA(", best_p, ", 1,", best_q, ")\n")
cat("Best AIC:", round(best_aic, 2), "\n")
summary(best_model)
```

### 9.3 Residual Diagnostics

```{r arima-diagnostics, fig.width=12, fig.height=8}
# A well-fitted model should have residuals that behave like white noise:
# - Random scatter around zero (no pattern)
# - ACF within confidence bands (no remaining autocorrelation)
# - Normal distribution (QQ plot on the line)
# - Ljung-Box p-values above 0.05 (no significant autocorrelation)
sarima(train_data$diff_sales, p = best_p, d = 1, q = best_q)
```

### 9.4 Forecast & Evaluation

```{r arima-forecast, fig.width=12, fig.height=5}
# Forecast on test set and compute error metrics
n_ahead        <- nrow(test_data)
arima_forecast <- forecast(best_model, h = n_ahead)
arima_pred     <- as.numeric(arima_forecast$mean)

# Error metrics
arima_rmse <- rmse(test_data$diff_sales, arima_pred)
arima_mae  <- mae(test_data$diff_sales, arima_pred)

cat("ARIMA Test RMSE:", round(arima_rmse, 2), "\n")
cat("ARIMA Test MAE: ", round(arima_mae, 2), "\n")

# Plot forecast vs actual
plot_df <- data.frame(
  date     = test_data$date,
  actual   = test_data$diff_sales,
  forecast = arima_pred,
  lower    = as.numeric(arima_forecast$lower[, 2]),
  upper    = as.numeric(arima_forecast$upper[, 2])
)

ggplot(plot_df, aes(x = date)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = actual),   color = "black",     lwd = 0.8) +
  geom_line(aes(y = forecast), color = "steelblue", lwd = 1) +
  labs(
    title    = paste0("ARIMA(", best_p, ",1,", best_q, ") Forecast vs Actual"),
    subtitle = "Black = Actual | Blue = Forecast | Shaded = 95% CI",
    x = "Date", y = "Differenced Sales"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

---

## 10. Model 2: SARIMA

### 10.1 Model Specification

The **SARIMA(p, d, q)(P, D, Q)[s]** model extends ARIMA with seasonal components:

$$\Phi_P(B^s)\phi_p(B)(1-B^s)^D(1-B)^d y_t = \Theta_Q(B^s)\theta_q(B)\epsilon_t$$

Where:
- $(p, d, q)$ = non-seasonal AR, differencing, MA orders
- $(P, D, Q)$ = seasonal AR, differencing, MA orders
- $s$ = seasonal period (7 for weekly data)
- $B^s$ = seasonal backshift operator: $B^s y_t = y_{t-s}$

**For our data:** $s = 7$ (weekly seasonality confirmed by ACF/PACF and decomposition).

```{r sarima-model, warning=FALSE, message=FALSE}
# SARIMA(1,1,1)(1,1,1)[7] — non-seasonal + weekly seasonal components
# Seasonal D=1 accounts for year-over-year weekly pattern differences
sarima_model <- tryCatch(
  auto.arima(
    ts(d_sales$total_sales[1:train_size], frequency = 7),
    seasonal      = TRUE,
    stepwise      = TRUE,
    approximation = TRUE,
    trace         = FALSE
  ),
  error = function(e) {
    # Fallback to manual specification if auto.arima fails
    Arima(ts(d_sales$total_sales[1:train_size], frequency = 7),
          order          = c(1, 1, 1),
          seasonal       = list(order = c(1, 1, 1), period = 7))
  }
)

cat("SARIMA model selected:\n")
print(sarima_model)
```

### 10.2 Residual Diagnostics

```{r sarima-diagnostics, fig.width=12, fig.height=8}
checkresiduals(sarima_model)
```

### 10.3 Forecast & Evaluation

```{r sarima-forecast, fig.width=12, fig.height=5}
sarima_forecast <- forecast(sarima_model, h = n_ahead)
sarima_pred     <- as.numeric(sarima_forecast$mean)

# Align length if needed
min_len      <- min(length(sarima_pred), nrow(test_data))
sarima_pred  <- sarima_pred[1:min_len]
actual_test  <- test_data$diff_sales[1:min_len]
test_dates   <- test_data$date[1:min_len]

sarima_rmse <- rmse(actual_test, sarima_pred)
sarima_mae  <- mae(actual_test, sarima_pred)

cat("SARIMA Test RMSE:", round(sarima_rmse, 2), "\n")
cat("SARIMA Test MAE: ", round(sarima_mae, 2), "\n")

plot_df2 <- data.frame(
  date     = test_dates,
  actual   = actual_test,
  forecast = sarima_pred,
  lower    = as.numeric(sarima_forecast$lower[1:min_len, 2]),
  upper    = as.numeric(sarima_forecast$upper[1:min_len, 2])
)

ggplot(plot_df2, aes(x = date)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "seagreen", alpha = 0.2) +
  geom_line(aes(y = actual),   color = "black",    lwd = 0.8) +
  geom_line(aes(y = forecast), color = "seagreen", lwd = 1) +
  labs(
    title    = "SARIMA Forecast vs Actual",
    subtitle = "Black = Actual | Green = Forecast | Shaded = 95% CI",
    x = "Date", y = "Differenced Sales"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

---

## 11. Model 3: TBATS

### 11.1 Model Specification

**TBATS** handles multiple seasonal periods simultaneously — ideal for retail data with both weekly (period = 7) and annual (period = 365.25) cycles.

The TBATS model is defined as:

$$y_t^{(\omega)} = l_{t-1} + \phi b_{t-1} + \sum_{i=1}^{T} s_t^{(i)} + d_t$$

Where:
- $y_t^{(\omega)}$ = Box-Cox transformed observations (stabilizes variance)
- $l_t$ = local level (trend intercept)
- $b_t$ = local slope (trend gradient), damped by $\phi$
- $s_t^{(i)}$ = $i$-th seasonal component modeled by trigonometric terms
- $d_t$ = ARMA($p$, $q$) error process

**Key advantage over ARIMA/SARIMA:** handles non-integer and multiple seasonal periods without manual specification. The model automatically selects the optimal Box-Cox $\lambda$, damping, and ARMA orders.

```{r tbats-model, warning=FALSE, message=FALSE}
# TBATS automatically detects seasonal periods and fits accordingly
# Use msts() to specify multiple seasonal periods: weekly (7) and annual (365.25)
sales_msts <- msts(
  d_sales$total_sales[1:train_size],
  seasonal.periods = c(7, 365.25)
)

tbats_model <- tbats(sales_msts)
cat("TBATS model summary:\n")
print(tbats_model)
```

### 11.2 Forecast & Evaluation

```{r tbats-forecast, fig.width=12, fig.height=5}
tbats_forecast <- forecast(tbats_model, h = n_ahead)
tbats_pred     <- as.numeric(tbats_forecast$mean)[1:min_len]

tbats_rmse <- rmse(actual_test, tbats_pred)
tbats_mae  <- mae(actual_test, tbats_pred)

cat("TBATS Test RMSE:", round(tbats_rmse, 2), "\n")
cat("TBATS Test MAE: ", round(tbats_mae, 2), "\n")

plot_df3 <- data.frame(
  date     = test_dates,
  actual   = actual_test,
  forecast = tbats_pred,
  lower    = as.numeric(tbats_forecast$lower[1:min_len, 2]),
  upper    = as.numeric(tbats_forecast$upper[1:min_len, 2])
)

ggplot(plot_df3, aes(x = date)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "darkorange", alpha = 0.2) +
  geom_line(aes(y = actual),   color = "black",      lwd = 0.8) +
  geom_line(aes(y = forecast), color = "darkorange",  lwd = 1) +
  labs(
    title    = "TBATS Forecast vs Actual",
    subtitle = "Black = Actual | Orange = Forecast | Shaded = 95% CI",
    x = "Date", y = "Sales"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

---

## 12. Model Comparison

```{r model-comparison}
comparison <- data.frame(
  Model = c(
    paste0("ARIMA(", best_p, ",1,", best_q, ")"),
    "SARIMA (auto-selected)",
    "TBATS (multi-seasonal)"
  ),
  RMSE = round(c(arima_rmse, sarima_rmse, tbats_rmse), 2),
  MAE  = round(c(arima_mae,  sarima_mae,  tbats_mae),  2),
  Handles_Seasonality  = c("No",  "Yes (single)", "Yes (multiple)"),
  Interpretability     = c("High", "High",         "Low"),
  Complexity           = c("Low",  "Medium",        "High")
)

knitr::kable(comparison, caption = "Model Comparison: ARIMA vs SARIMA vs TBATS")
```

---

## 13. Future Forecast (30 Days)

```{r future-forecast, fig.width=12, fig.height=6}
# Use the best model (lowest RMSE) to forecast 30 days beyond the training period
# Here we use SARIMA as it balances accuracy and interpretability

future_forecast <- forecast(sarima_model, h = 30)

autoplot(future_forecast) +
  labs(
    title    = "30-Day Sales Forecast (SARIMA)",
    subtitle = "Shaded regions = 80% and 95% prediction intervals",
    x        = "Time",
    y        = "Total Sales"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

---

## 14. Conclusions

```{r conclusions, echo=FALSE}
cat("
ANALYSIS SUMMARY
================

Data Findings:
- Strong weekly seasonality confirmed (7-day cycle) — weekends consistently outsell weekdays
- Mild upward trend through 2014-2015
- Christmas Day is a structural zero — stores closed; treated as outlier and removed
- First-order differencing is sufficient to achieve stationarity (ADF test confirmed)

Model Results:
- ARIMA: Fastest to fit, interpretable, but ignores seasonality in its standard form
- SARIMA: Best balance of accuracy and interpretability; auto.arima selected
  optimal seasonal orders; handles weekly cycle explicitly
- TBATS: Handles both weekly and annual seasonality simultaneously;
  useful when longer history is available to identify annual patterns

Recommendation:
- For operational forecasting (inventory planning), SARIMA is recommended —
  it captures the dominant weekly pattern and provides interpretable coefficients
- TBATS becomes more valuable with 3+ years of data where annual seasonality is clear
- ARIMA alone is insufficient for retail data with strong weekly patterns

Limitations & Next Steps:
1. Aggregate forecasting hides item-level and store-level variation —
   hierarchical forecasting would improve operational utility
2. External features (events, SNAP days, holidays) from the calendar file
   were not incorporated — adding them as regressors (ARIMAX) would improve accuracy
3. LightGBM or gradient boosting (as used by M5 competition winners) significantly
   outperforms classical time series methods on this dataset
4. Prophet (Facebook) handles holidays and seasonality automatically and
   would be a strong next model to test
")
```

---

## 15. Methodology Notes

**Why first-order differencing?**
Differencing removes trend by computing $\Delta y_t = y_t - y_{t-1}$. We apply it once ($d=1$) — confirmed sufficient by ADF test on the differenced series.

**Why AIC for model selection?**
$$\text{AIC} = 2k - 2\ln(\hat{L})$$
AIC penalizes model complexity ($k$ = number of parameters) to prevent overfitting while rewarding goodness of fit ($\hat{L}$ = likelihood). Lower AIC = better model.

**Why RMSE and MAE?**
$$\text{RMSE} = \sqrt{\frac{1}{n}\sum_{t=1}^{n}(y_t - \hat{y}_t)^2}$$
$$\text{MAE} = \frac{1}{n}\sum_{t=1}^{n}|y_t - \hat{y}_t|$$
RMSE penalizes large errors more heavily (useful when big misses are costly). MAE treats all errors equally. Together they give a complete picture of forecast accuracy.
